---
title: "The most severe weather events across the US from 1950 to 2011"
author: "Fernando Barranco R"
date: "30 de julio de 2017"
output: html_document
---

# Course Project 2

## The most severe weather events across the US from 1950 to 2011

*Important: If you want to run the entire code, just get rid of all the '#' symbols whenever you see an R chunk.*

Synopsis: Severe weather events can cause both public health (fatalities, injuries) and economic problems (property damage) for communities, and preventing such outcomes is a key concern; this project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database wich tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage, the events in the database start in the year 1950 and end in November 2011.
  
The goal of the proyect is to answer two questions; across the United States:
  
* Which types of events are most harmful with respect to population health?
  
* Which types of events have the greatest economic consequences?
  
Set working directory

```{r}
setwd("C:/Users/FernandoBarranco/Desktop")
```

### Data Processing

Download the data from the following url (if necessary)

```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
# download.file(url, "C:/Users/FernandoBarranco/Desktop/StormData.csv.bz2")
```

Unzip the .csv file

```{r}
# install.packages("R.utils")
# library(R.utils)
# bunzip2("StormData.csv.bz2", "StormData.csv", remove = FALSE, skip = TRUE)
```

Read the data into R

```{r}
# install.packages("data.table")
library(data.table)
storm_data <- fread("StormData.csv")
```

### Subset, transform and clean the data

Rename variables

```{r}
# install.packages("dplyr")
library(dplyr)
# install.packages("lubridate")
library(lubridate)
variables <- names(storm_data)
names(storm_data) <- tolower(variables)
```

Keep variables to analyze 

```{r}
storm_data_sub <- storm_data %>% select(bgn_date, state, evtype, fatalities,
                      injuries, propdmg, propdmgexp, cropdmg, cropdmgexp)
```

Change to date format

```{r}
storm_data_sub$bgn_date <- as.Date(storm_data_sub$bgn_date, format = "%m/%d/%Y %H:%M:%S")
```

Removing observations with no fatalities nor injuries

```{r}
storm_data_sub <- storm_data_sub %>%
        mutate(year = year(bgn_date), evtype = tolower(evtype)) %>%
        select(-bgn_date, state, evtype, fatalities, injuries, year) %>%
        filter(fatalities != 0 | injuries != 0)
head(storm_data_sub)
```

### Data wrangling and feature engineering

Sum up fatalities and injuries by event type

```{r}
fatalities <- aggregate(fatalities ~ evtype, storm_data_sub, sum)
head(fatalities)
injuries <- aggregate(injuries ~ evtype, storm_data_sub, sum)
head(injuries)
```

Keep extreme values that causes the highest damages

```{r}
storm_data_health <- inner_join(fatalities, injuries, by = "evtype") %>%
        filter(fatalities > quantile(fatalities, probs = .97) &
                injuries > quantile(injuries, probs = .97))
storm_data_health
```
  
*Note: The 'storm_data_health' dataset will help us answer question 1*
  
Property damages counts

```{r}
table(storm_data_sub$propdmgexp)
table(storm_data_sub$cropdmgexp)
```

Map property damage alphanumeric exponents to numeric values.

```{r}
propdmgkey_level <- c("-", "0", "5", "7", "H", "K", "M", "B")
propdmgkey_label <- c( 10^0+1 , 10^0, 10^5, 10^7, 10^2, 10^3, 10^6, 10^9 )
```

Map crop damage alphanumeric exponents to numeric values

```{r}
cropdmgkey_level <- c("0", "K", "M", "B")
cropdmgkey_label <- c( 10^0, 10^3, 10^6, 10^9 )
```

Create new variables that contain the total property damage

```{r}
storm_data_eco <- storm_data_sub %>%
        select(-(c(state, fatalities, injuries, year))) %>%
        mutate(propdmgexp = toupper(propdmgexp),
               cropdmgkey = factor(cropdmgexp, cropdmgkey_level, cropdmgkey_label),
               propdmgkey = factor(propdmgexp, propdmgkey_level, propdmgkey_label),
               cropdmgkey = as.numeric(as.character(cropdmgkey)),
               propdmgkey = as.numeric(as.character(propdmgkey)),
               propdmgtot = propdmg * propdmgkey,
               cropdmgtot = cropdmg * cropdmgkey )
head(storm_data_eco)
```

Sum up property damage by event type

```{r}
propdmgtotal <- aggregate(propdmgtot ~ evtype, storm_data_eco, sum)
head(propdmgtotal)
cropdmgtotal <- aggregate(cropdmgtot ~ evtype, storm_data_eco, sum)
head(cropdmgtotal)
```

Keep extreme values that causes the highest damages

```{r}
storm_data_economy <- full_join(propdmgtotal, cropdmgtotal, by = "evtype") %>%
        filter( propdmgtot > quantile(propdmgtot, probs = .90, na.rm = T) &
                cropdmgtot > quantile(cropdmgtot, probs = .90, na.rm = T))
storm_data_economy
```
  
*Note: The 'storm_data_economy' dataset will help us answer question 2*
  
### Results

```{r}
# install.packages("ggplot2")
library(ggplot2)
```

* Which types of events are most harmful with respect to population health?

```{r}
ggplot(storm_data_health, aes(injuries, fatalities, color = evtype, shape = evtype)) +
        geom_point(size = 8, alpha = 1) +
        guides(color = guide_legend("Event")) +
        theme_set(theme_bw()) + 
        labs(x = "Injuries", y = "Fatalities",
             title = "The most harmful Events with respect to population Health") +
        scale_shape_manual(name = "Event",
                           values = c(15, 16, 17, 20, 19, 18))
```

Notice that tornado has the highest values for both injuries and fatalities, this event
is the most harmful to population health.

* Which types of events have the greatest economic consequences?

```{r}
ggplot(storm_data_economy, aes(propdmgtot/10^9, cropdmgtot/10^9, shape = evtype)) +
        geom_point(size = 8, alpha = 1, aes(color = evtype)) +
        guides(color = guide_legend("Event")) +
        theme_set(theme_bw()) + 
        labs(x = "Property Damage (Billions $)", y = "Crop Damage (Billions $)",
             title = "Events with the greatest Economic consequences") +
        scale_shape_manual(name = "Event",
                           values = c(15, 16, 17, 20, 19, 18))
```

Notice that hurricane/typhoon has the highest values for property damage, this event
causes the greatest economic losses.